Tackling discrepancies in freshwater invertebrate trait databases: Harmonising across continents and aggregating taxonomic resolution

### 2021-06-28 Paper/Submission_new/Main_Document-Latex.tex
21.02: Finalised files for submission
28.06.2021: Update paper revisions

\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{xr-hyper}
\usepackage{hyperref}
\usepackage{float}
\usepackage[table,xcdraw]{xcolor}
\usepackage{color, colortbl}
\usepackage{longtable}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{tikz}
\usepackage{rotating}
\usepackage{caption}
\usepackage{authblk}
\usepackage{csquotes}
\graphicspath{{Figures/}}
\usepackage{setspace}
\usepackage{rotating}
\usepackage{geometry}
\usepackage{array}
\usepackage{lscape}
\usepackage{longtable}
\usepackage{etoolbox}
\usepackage{hhline}
\usepackage{lmodern}
\usepackage[
backend = biber,
natbib,
citestyle = authoryear,
bibstyle = apa6,
maxcitenames = 2, 
maxbibnames = 99, 
uniquename=false,
uniquelist=false
%sorting = none % sort by name year title
]{biblatex}
\addbibresource{Ref_invertebrate_DB.bib}

@@

\myexternaldocument{SI_test}

@@

4. By identifying discrepancies in trait definitions we hope to motivate the development of standardised terminology for invertebrate traits. Our results also illustrate the usefulness of harmonised datasets for ecological study and provide guidance for the circumstances under which the choice of trait aggregation method is important.
\\
\keywords{trait definitions, trait aggregation, data synthesis}

@@

Invertebrate traits have been increasingly used in freshwater ecology, e.g. by relating macroinvertebrate trait composition to environmental factors such as climate change or salinisation (\cite{bhowmik_large_2015, poff_developing_2010, szocs_effects_2014}). In the last decades, freshwater ecologists have compiled comprehensive invertebrate trait databases, which typically include data for a single region or continent (\cite{kefford_integrated_2020, Philips_and_Smith_NZ_DB_2018, schmidt-kloiber_www.freshwaterecology.info_2015, tomanova_trophic_2006, ussegliopolatera_biological_2000, vieira_database_nodate}). The availability of invertebrate trait data from different continents enables comparisons of regional trait variation and allows for testing the consistency of trait structure–environmental factor relationships across both small and large spatial extents. To date, such analyses have been carried out mostly within continents, using information from one or two trait databases. For example, \citet{bonada_taxonomic_2007} compared trait composition for Mediterranean and temperate regions in Europe based on traits from \citet{ussegliopolatera_biological_2000} (typically referred to as the Tachet database), \citet{poff_developing_2010} characterised trait composition across sites in the Western USA based on traits from \citet{poff_functional_2006}, and \citet{botwe_effects_2018} used trait definitions from \citet{poff_functional_2006} and \citet{schafer_trait_2011} to test for effects of salinity on invertebrate traits across different sites in South Australia. Analyses that synthesise invertebrate trait information from more than two different continents are rare, but see \citet{brown_functional_2018} and \citet{statzner_reproductive_1997}.

@@

A third challenge to comparing across invertebrate trait databases is inconsistent coding of trait data (\cite{culp_incorporating_2011}). Traits of individual freshwater invertebrates can be difficult to quantify because, unlike plant traits, they are often difficult to measure directly. For example, to describe feeding habits requires that we understand mouthpart morphology, feeding behaviour, and what is being eaten (\cite{moog_comprehensive_nodate}). Some traits can be described in categorical terms, which ignores uncertainty in how traits are expressed in any particular organism (e.g., adult terrestrial stage, presence of gills). Other traits are better suited to being described as continuous (e.g., tolerance of pollution, body size). One approach for dealing with uncertainty is the use of fuzzy coding, where traits are assigned probabilistic values. Fuzzy codes, which are usually converted to proportions, are used to account for plasticity in traits, variability in traits within taxonomic groups above the species level, and incomplete knowledge. The values derived from the different coding approaches represent affinities (aka affinity scores) that indicate how strongly a taxon expresses a particular trait. The use of different types of coding for trait affinities in different databases limits comparisons. For example, in the study from \citet{brown_functional_2018} the authors needed to reclassify trait affinities because the individual databases employed different coding approaches (i.e., European and New Zealand databases used fuzzy coding whereas the North American database used binary coding).

@@

We established four datasets, one for each of the four geographic regions, by harmonising the traits of seven grouping features from the trait databases. We selected traits that were available in all databases, are commonly applied in trait-based ecological studies, and describe different aspects of the biology of a species: life history (Voltinism), morphology (Respiration, Body form, Size), ecology (Locomotion, Feeding mode) and reproduction (Oviposition). We omitted ecological traits that describe habitat preferences (e.g. temperature preference) because these traits are missing in the New Zealand trait database. Grouping features were classified differently across the databases, so we harmonised 7 grouping features that collectively included 26 traits (Table \ref{tab:traits_harmonisation}). To do so we combined similar traits into one trait (e.g. crawlers and sprawlers into crawlers) by assigning for each taxon for the new combined trait the highest affinity score of those originally assigned. Next, we selected taxa for inclusion in our datasets and normalised the trait affinity scores for each taxon. We first omitted all taxa with a lower taxonomic precision than family level and consolidated duplicate taxa that appeared within a database, by either applying the median for fuzzy coded traits, or the maximum for binary traits. We then used either fuzzy coding or, when unavailable, binary coding to create the final harmonised datasets. Fuzzy codes are reported with different ranges in the trait databases (e.g., freshwaterecology.info: 0–10, Tachet database: 0–3 or 0–5). We normalised these to a range between 0 and 1 and converted trait affinities to proportions. In cases where binary coding was needed, we converted categorical and continuous traits into binary traits with values of 1 and 0, which, therefore, had the same range as fuzzy coded traits. For example, in the  freshwaterecology.info database, voltinism traits are associated with different faunistic regions, and we substituted a value of 1 for entries such as “arctic” or “boreal”. We assumed that values of 1 and 0 for binary traits corresponded to the highest and lowest affinity for a particular trait.

@@

To compare the effects of different aggregation methods on trait\-environment relationship inferences, we aggregated the traits of two of the harmonised datasets to the family level. We used three different aggregation approaches: (1) direct aggregation of taxa to family level, in which we aggregated either trait affinities by the median or mean for species and genera within a family, giving equal weight to every taxon. We denote these aggregation methods as \textit{direct\_agg\textsubscript{median}} and \textit{direct\_agg\textsubscript{mean}}, respectively; and (2) stepwise aggregation, first to the genus level and subsequently to the family level with either the median or mean. This approach gives equal weights to each genus. Hereafter, we denote this aggregation type as \textit{stepwise\_agg\textsubscript{median}} or \textit{stepwise\_agg\textsubscript{mean}}, respectively; and (3) aggregation using a weighted mean approach, denoted as \textit{weighted\_agg}. This method weights each genus according to its number of species in the trait datasets, regardless of whether information was available for every used grouping feature (Figure \ref{fig:data_proc_overview}). When we refer to \textit{direct\_agg\textsubscript{mean}}, \textit{stepwise\_agg\textsubscript{mean}}, and \textit{weighted\_agg} together we denote these methods as mean aggregation methods and when we refer to \textit{direct\_agg\textsubscript{median}} and \textit{stepwise\_agg\textsubscript{median}} together, we denote these methods as median aggregation methods.

@@

\subsubsection*{Effect of phylogeny and trait variability on aggregation outcomes}

To examine the influence of phylogeny and trait variability on the outcomes of the different trait aggregation methods, we created three hypothetical scenarios. We simulated three different families, each containing 25 total species but with different phylogenetic structures. The three families consisted of (1) a family with an equal number of genera and species (five genera each with five species each), denoted as \textit{sim\_base}; (2) a family in which one genus had a much larger number of species than the other four genera (1 genus with 13 species, 4 genera with 3 species each), denoted as \textit{sim\_extreme}; (3) a family in which all genera had a different number of species (8, 2, 7, 3, 5), denoted as \textit{sim\_variation}. We assigned a hypothetical grouping feature with three traits (T1, T2, and T3) to each scenario. We then simulated the 25 affinities, one per species, for each trait, by sampling from a truncated normal distribution bound by 0 and 1 and with a mean value of 0.5. To simulate different levels of trait variability, we repeated the sampling 100 times for each of the 5 standard deviations (0.2, 0.4, 0.6, 0.8, and 1), resulting in 12,500 simulated trait affinities for each simulated trait ($25$ species per family $\times$ 5 levels of trait variability $\times$ 100 replicates). We converted simulated trait affinities to proportions, as was done during trait database processing, and assigned the 12,500 simulated affinities per trait to each of the three family scenarios. We then applied the five trait aggregation methods described above to each simulated dataset. We compared the resulting ranges of aggregated trait affinities between levels of trait variability and phylogenetic scenario as well as the differences in trait affinities obtained by each aggregation method.

@@

We re-analysed the \citet{szocs_effects_2014} data by performing a RDA with harmonised grouping features from the dataset for Europe and with traits that we aggregated to family level with the five aggregation methods described above. We used all 21 grouping features from the original study, but replaced 6 with harmonised grouping features from the established European trait dataset, namely size, feeding mode, locomotion, oviposition, respiration, and voltinism.
We compared the results of the original study’s RDA with the results of the RDA from the harmonised grouping features. Specifically, trait composition, expressed as community weighted mean (CWM) traits, was ordinated along an electrical conductivity gradient. We compared the species scores obtained from the RDA (i.e., the coordinates of the tips of the vectors representing the CWM traits in the bi- or triplots). Following the original study, we used the Mahalanobis distance measure and identified traits associated with higher or lower salinity based on their distance to the ordination axis median. We considered CWM traits with a Mahalanobis distance greater than the 97.5 \%-quantile of the Chi-square distribution (5.02) to have responded to either lower or higher salinity. Additionally, to test for the effect of different aggregation methods on inferences about CWM trait–salinisation relationships, we repeated the RDA five times, each with family-level aggregated trait values assigned to corresponding taxa from \citet{szocs_effects_2014}.

@@

The definitions of grouping features also varied across databases in their differentiation into traits and in their focus, both which can lead to discrepancies in trait definitions (see Table \ref{stab:trait_definitions} for a summary of discrepancies in trait definitions). Varying levels of differentiation among the trait databases were present in all investigated grouping features (Table \ref{tab:trait_databases_coding_differentiation} and Table \ref{stab:trait_definitions}).

@@

\subsubsection*{Comparison of aggregation methods with varying phylogenies and trait variability}

The simulations showed that both phylogenetic structure and trait variability affected aggregated trait affinities, although only to a small degree, and that aggregation method mattered in terms of the ranges of trait affinities produced over simulation replicates. The effect of phylogenetic structure differed across aggregation methods, but, as we expected, the range of trait affinities increased with increasing trait variability for all aggregation methods (Figure \ref{fig:overview_sim_results}).

Phylogenetic structure appeared to influence the outcomes of the different aggregation methods. For the \textit{sim\_base} scenario (equal numbers of genera and species), the mean and weighted aggregation methods yielded similar ranges of aggregated trait affinities within each level of trait variability, and the median aggregation methods consistently produced greater ranges of aggregated trait affinities than the other methods. The largest ranges were produced by the \textit{stepwise\_agg \textsubscript{median}} method. For the more complex phylogenetic structures, \textit{sim\_extreme} (one genus with a much larger number of species than the other four) and \textit{sim\_variation} (all genera with a different number of species), the \textit{stepwise\_agg \textsubscript{median}} method still produced the largest ranges of trait affinities for most levels of trait variability (and \textit{direct\_agg\textsubscript{mean}} produced the narrowest ranges for most levels of trait variability), but there was much less consistency in the ranges of trait affinities for all aggregation methods (i.e., the ranges were different among all aggregation methods).

Although trait aggregation methods were affected by phylogenetic structure and trait variability to some degree, in most simulated datasets, the different aggregation methods resulted in similar trait affinities. Only 1.4 \%, or 213 out of 15.000 total comparisons (3 scenarios $\times$ 5 levels of trait variability $\times$ 10 unique comparisons of trait aggregation methods $\times$ 100 replicates) showed a difference equal or greater than an absolute trait affinity of 0.1. Most  (83.5 \%) of these differences occurred in the \textit{sim\_extreme} scenario and were found between the aggregation methods \textit{direct\_agg\textsubscript{mean}} and \textit{stepwise\_agg\textsubscript{median}}, \textit{direct\_agg\textsubscript{median}} and \textit{stepwise\_agg\textsubscript{median}}, and \textit{stepwise\_agg\textsubscript{median}} and \textit{weighted\_agg} (Figure \ref{fig:sim_indv_runs}).

\begin{figure}[H]
  \centering
  \includegraphics[width=16.5cm, height=10cm]{Overview_sim_results.png}
  \caption{Ranges of aggregated trait affinities for the three examples of taxonomic hierarchies and simulated levels of trait variability. Shown are the results only for one simulated trait (T1). Similar results were obtained for the other simulated traits (Figure \ref{fig:overview_sim_results_T2_T3}). Boxplots depict results for 100 replicated simulations of each trait aggregation method. The boxplot depicts the median and encompasses the 25\textsuperscript{th} and 75\textsuperscript{th} percentile. Horizontal black lines depict the median. Whiskers extend to the largest and smallest value respectively no further than 1.5 $\times$ the inter-quartile range. Outliers beyond the end of the whiskers are plotted as grey dots. Trait aggregation methods are in order of least to greatest produced ranges to improve visual inspection.}
  \label{fig:overview_sim_results}
\end{figure}

\begin{figure}[H]
  \centering
  \includegraphics[width=16.5cm, height=10cm]{Diffs_indiv_runs_sim.png}
  \caption{Comparison between the aggregated trait affinities produced by the different trait aggregation methods for every simulated dataset across all 3 simulated traits (5,000 comparisons per scenario). Results are shown for different levels of trait variability which is represented by the standard deviation (SD) of the simulated traits. Dots depict comparisons where absolute differences between aggregated trait affinities were greater than 0.1. Overall, there are 10 possible unique comparisons of which 7 produced absolute differences greater than 0.1. \newline
  Comparisons: \newline
  1) \textit{direct\_agg (median)} - \textit{stepwise\_agg (mean)} \newline
  2) \textit{direct\_agg (median)} - \textit{weighted\_agg} \newline
  3) \textit{stepwise\_agg (mean)} - \textit{stepwise\_agg (median)} \newline
  4) \textit{stepwise\_agg (mean)} - \textit{weighted\_agg} \newline  
  5) \textit{direct\_agg (mean)} - \textit{stepwise\_agg (median)} \newline
  6) \textit{direct\_agg (median)} - \textit{stepwise\_agg (median)} \newline
  7) \textit{stepwise\_agg (median)} - \textit{weighted\_agg} \newline
  }
  \label{fig:sim_indv_runs}
\end{figure}

@@

\section*{Acknowledgments}

@@

% Table 1

@@

% Table 5
\begin{landscape}
\begin{table}[H]
  \centering
  \caption{Percentage of differing cases, minimum, maximum, mean, and standard deviation of absolute differences between trait affinities assigned at family level by experts and aggregated trait affinities from five different aggregation methods.}
  \label{tab:summary_stat_aggr_vs_fam_assigned}
  \begin{tabular}{ll|ccccc}
  \toprule[.1em]
  \specialcell{Data \\ origin} & \specialcell{Comparison to\\ traits at family level} & \specialcell{Differing \\ cases [\%]} & \specialcell{Min. \\ differences} & \specialcell{Max. \\ differences} & \specialcell{Mean abs. \\ differences} & \specialcell{SD abs. \\ differences} \\ 
  \toprule[.1em]
  \multirow{4}{*}{\specialcell{AUS}} & \specialcell{\textit{direct\_agg (median)}} & 16.53 & 0.01 & 1.00 & 0.45 & 0.27 \\ 
  & \specialcell{\textit{direct\_agg (mean)}} & 23.24 & $< 0.01$ & 0.99 & 0.34 & 0.23 \\ 
  & \specialcell{\textit{stepwise\_agg (median)}} & 17.90 & 0.01 & 1.00 & 0.42 & 0.26 \\ 
  & \specialcell{\textit{stepwise\_agg (mean)}} & 23.24 & $< 0.01$ & 0.99 & 0.33 & 0.22 \\ 
  & \specialcell{\textit{weighted\_agg}} & 23.24 & $< 0.01$ & 1.00 & 0.34 & 0.24 \\ 
  \midrule
  \multirow{4}{*}{\specialcell{NA}} & \specialcell{\textit{direct\_agg (median)}} & 15.33 & 0.17 & 1.00 & 0.70 & 0.26 \\ 
  & \specialcell{\textit{direct\_agg (mean)}} & 47.00 & $< 0.01$ & 1.00 & 0.30 & 0.26 \\ 
  & \specialcell{\textit{stepwise\_agg (median)}} & 18.00 & 0.08 & 1.00 & 0.63 & 0.28 \\ 
  & \specialcell{\textit{stepwise\_agg (mean)}} & 47.00 & $< 0.01$ & 1.00 & 0.30 & 0.27 \\ 
  & \specialcell{\textit{weighted\_agg}} & 47.00 & $< 0.01$ & 1.00 & 0.31 & 0.28 \\ 
  \bottomrule
  \end{tabular}
\end{table}
\begin{minipage}{\linewidth}{\fontsize{8}{10}\selectfont
  \centering
  Abbreviations: Min., Minimum; Max., Maximum; abs., absolute; SD, Standard deviation; AUS, Australia; NA, North America.
  }
 \end{minipage}
\end{landscape}

@@

Ranges of aggregated trait affinities for the three examples of taxonomic hierarchies and simulated levels of trait variability. Shown are the results only for one simulated trait (T1). Similar results were obtained for the other simulated traits (Figure \ref{fig:overview_sim_results_T2_T3}). Boxplots depict results for 100 replicated simulations of each trait aggregation method. The boxplot depicts the median and encompasses the 25\textsuperscript{th} and 75\textsuperscript{th} percentile. Horizontal black lines depict the median. Whiskers extend to the largest and smallest value respectively no further than 1.5 $\times$ the inter-quartile range. Outliers beyond the end of the whiskers are plotted as grey dots. Trait aggregation methods are in order of least to greatest produced ranges to improve visual inspection.

@@

Comparison between the aggregated trait affinities produced by the different trait aggregation methods for every simulated dataset across all 3 simulated traits (5,000 comparisons per scenario). Results are shown for different levels of trait variability which is represented by the standard deviation (SD) of the simulated traits. Dots depict comparisons where absolute differences between aggregated trait affinities were greater than 0.1. Overall, there are 10 possible unique comparisons of which 7 produced absolute differences greater than 0.1. \newline
  Comparisons: \newline
  1) \textit{direct\_agg (median)} - \textit{stepwise\_agg (mean)} \newline
  2) \textit{direct\_agg (median)} - \textit{weighted\_agg} \newline
  3) \textit{stepwise\_agg (mean)} - \textit{stepwise\_agg (median)} \newline
  4) \textit{stepwise\_agg (mean)} - \textit{weighted\_agg} \newline  
  5) \textit{direct\_agg (mean)} - \textit{stepwise\_agg (median)} \newline
  6) \textit{direct\_agg (median)} - \textit{stepwise\_agg (median)} \newline
  7) \textit{stepwise\_agg (median)} - \textit{weighted\_agg} \newline

>>>>

\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{xr-hyper}
\usepackage{hyperref}
\usepackage{float}
\usepackage[table,xcdraw]{xcolor}
\usepackage{color, colortbl}
\usepackage{longtable}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{tikz}
\usepackage{rotating}
\usepackage{caption}
\usepackage{authblk}
\usepackage{csquotes}
\graphicspath{{Figures/}}
\usepackage{setspace}
\usepackage{rotating}
\usepackage{geometry}
\usepackage{array}
\usepackage{lscape}
\usepackage{longtable}
\usepackage{etoolbox}
\usepackage{hhline}
\usepackage{lmodern}
\usepackage[
backend = biber,
natbib,
citestyle = authoryear,
bibstyle = apa,
maxcitenames = 2, 
maxbibnames = 99, 
uniquename=false,
uniquelist=false
%sorting = none % sort by name year title
]{biblatex}
\addbibresource{Ref_invertebrate_DB.bib}

@@

\myexternaldocument{Latex_Supplementary_File}

@@

4. By identifying discrepancies in trait definitions we hope to motivate the development of standardised terminology for invertebrate traits. Our results also illustrate the usefulness of harmonised datasets for ecological study and provide guidance for the circumstances under which the choice of trait aggregation method is important.
\\
\keywords{trait definitions, trait aggregation, data synthesis, trait-environment relationships, large-scale comparisons}

@@

Invertebrate traits have been increasingly used in freshwater ecology, e.g. by relating macroinvertebrate trait composition to environmental factors such as climate change or salinisation (\cite{bhowmik_large_2015, poff_developing_2010, szocs_effects_2014}), for predicting changes in ecosystem functioning and trophic dynamics (\cite{vos_taxonomic_2017, gutierrez-canovasPopulationsHighvaluePredators2021}), or for developing indices for biomonitoring (\cite{beketov_spear_2009}). In the last decades, freshwater ecologists have compiled comprehensive invertebrate trait databases, which typically include data for a single region or continent (\cite{kefford_integrated_2020, Philips_and_Smith_NZ_DB_2018, schmidt-kloiber_www.freshwaterecology.info_2015, tomanova_trophic_2006, ussegliopolatera_biological_2000, vieira_database_nodate, gayraudInvertebrateTraitsBiomonitoring2003, statznerConservationTaxonomicBiological2007}). The availability of invertebrate trait data from different continents enables comparisons of regional trait variation and allows for testing the consistency of trait structure–environmental factor relationships across both small and large spatial extents. To date, such analyses have been carried out mostly within continents, using information from one or two trait databases. For example, \citet{bonada_taxonomic_2007} compared trait composition for Mediterranean and temperate regions in Europe based on traits from \citet{ussegliopolatera_biological_2000} (typically referred to as the Tachet database), \citet{poff_developing_2010} characterised trait composition across sites in the Western USA based on traits from \citet{poff_functional_2006}, and \citet{botwe_effects_2018} used trait definitions from \citet{poff_functional_2006} and \citet{schafer_trait_2011} to test for effects of salinity on invertebrate traits across different sites in South Australia. Analyses that synthesise invertebrate trait information from more than two different continents are rare, but see \citet{brown_functional_2018} and \citet{statzner_reproductive_1997}.

@@

A third challenge to comparing across invertebrate trait databases is inconsistent coding of trait data (\cite{culp_incorporating_2011}). Traits of individual freshwater invertebrates can be difficult to quantify because, unlike plant traits, they are often difficult to measure directly. For example, to describe feeding habits requires that we understand mouthpart morphology, feeding behaviour, and what is being eaten (\cite{moog_comprehensive_nodate}). Some traits can be described in categorical terms, which ignores uncertainty in how traits are expressed in any particular organism (e.g., adult terrestrial stage, presence of gills). Other traits are better suited to being described as continuous (e.g., tolerance of pollution, body size). One approach for dealing with uncertainty is the use of fuzzy coding, where traits are assigned probabilistic values (\cite{chevenet_francois_fuzzy_1994}). Fuzzy codes, which are usually converted to proportions, are used to account for plasticity in traits, variability in traits within taxonomic groups above the species level, and incomplete knowledge. The values derived from the different coding approaches represent affinities (aka affinity scores) that indicate how strongly a taxon expresses a particular trait. The use of different types of coding for trait affinities in different databases limits comparisons. For example, in the study from \citet{brown_functional_2018} the authors needed to reclassify trait affinities because the individual databases employed different coding approaches (i.e., European and New Zealand databases used fuzzy coding whereas the North American database used binary coding).

@@

We established four datasets, one for each of the four geographic regions, by harmonising the traits of seven grouping features from the trait databases. We selected traits that are commonly applied in trait-based ecological studies and describe different aspects of the biology of a species: life history (Voltinism), morphology (Respiration, Body form, Size), ecology (Locomotion, Feeding mode) and reproduction (Oviposition). The selected traits also relate to how taxa cope with disturbances (e.g. reproduction traits like ovoviviparity or number of life cycles) and to ecosystem functions such as organic matter breakdown and nutrient cycling (e.g. trait shredder). A criterion for trait selection was that the traits occurred in all databases, to enable us to show trait definitions differences and the influence of grouping-feature harmonisation on trait-environment relationships. Thus, we omitted ecological traits that describe habitat preferences (e.g. temperature preference) because these traits are missing in the New Zealand trait database.
% OLD:
% We selected traits that were available in all databases, are commonly applied in trait-based ecological studies, and describe different aspects of the biology of a species: life history (Voltinism), morphology (Respiration, Body form, Size), ecology (Locomotion, Feeding mode) and reproduction (Oviposition).

Grouping features were classified differently across the databases, so we harmonised 7 grouping features that collectively included 26 traits (Table \ref{tab:traits_harmonisation}). To do so we combined similar traits into one trait (e.g. crawlers and sprawlers into crawlers) by assigning for each taxon for the new combined trait the highest affinity score of those originally assigned. Next, we selected taxa for inclusion in our datasets and normalised the trait affinity scores for each taxon. We first omitted all taxa with a lower taxonomic precision than family level and consolidated duplicate taxa that appeared within a database, by either applying the median for fuzzy coded traits, or the maximum for binary traits. We then used either fuzzy coding or, when unavailable, binary coding to create the final harmonised datasets. Fuzzy codes are reported with different ranges in the trait databases (e.g., freshwaterecology.info: 0–10, Tachet database: 0–3 or 0–5). We normalised these to a range between 0 and 1 and converted trait affinities to proportions. In cases where binary coding was needed, we converted categorical and continuous traits into binary traits with values of 1 and 0, which, therefore, had the same range as fuzzy coded traits. For example, in the  freshwaterecology.info database, voltinism traits are associated with different faunistic regions, and we substituted a value of 1 for entries such as “arctic” or “boreal”. We assumed that values of 1 and 0 for binary traits corresponded to the highest and lowest affinity for a particular trait.

@@

To compare the effects of different aggregation methods on trait-environment relationship inferences, we aggregated the traits of two of the harmonised datasets to the family level. We used three different aggregation approaches: (1) direct aggregation of taxa to family level, in which we aggregated either trait affinities by the median or mean for species and genera within a family, giving equal weight to every taxon. We denote these aggregation methods as \textit{direct\_agg\textsubscript{median}} and \textit{direct\_agg\textsubscript{mean}}, respectively; and (2) stepwise aggregation, first to the genus level and subsequently to the family level with either the median or mean. This approach gives equal weights to each genus. Hereafter, we denote this aggregation type as \textit{stepwise\_agg\textsubscript{median}} or \textit{stepwise\_agg\textsubscript{mean}}, respectively; and (3) aggregation using a weighted mean approach, denoted as \textit{weighted\_agg}. This method weights each genus according to its number of species in the trait datasets, regardless of whether information was available for every used grouping feature (Figure \ref{fig:data_proc_overview}). When we refer to \textit{direct\_agg\textsubscript{mean}}, \textit{stepwise\_agg\textsubscript{mean}}, and \textit{weighted\_agg} together we denote these methods as mean aggregation methods and when we refer to \textit{direct\_agg\textsubscript{median}} and \textit{stepwise\_agg\textsubscript{median}} together, we denote these methods as median aggregation methods.

@@

Further, we examined the influence of phylogeny and trait variability on the outcomes of the different trait aggregation methods by simulating families with different phylogenetic structures and different levels of trait variability. The simulation showed that both phylogenetic structure and trait variability can affect aggregated trait affinities, but mostly only to a small degree. As expected, the range of trait affinities increasing trait variability for all trait aggregation methods. Therefore, we provide details of the simulation and the results in the supplementary materials.

@@

We re-analysed the \citet{szocs_effects_2014} data by performing a RDA with harmonised grouping features from the dataset for Europe and with traits that we aggregated to family level with the five aggregation methods described above. We used all 21 grouping features from the original study, but replaced 6 with harmonised grouping features from the established European trait dataset, namely size, feeding mode, locomotion, oviposition, respiration, and voltinism.
We compared the results of the original study’s RDA with the results of the RDA from the harmonised grouping features. Specifically, trait composition, expressed as community weighted mean (CWM) traits, was ordinated along an electrical conductivity gradient. We compared the species scores obtained from the RDA (i.e., the coordinates of the tips of the vectors representing the CWM traits in the bi- or triplots). Following the original study, we used the Mahalanobis distance measure and identified traits associated with higher or lower salinity based on their distance to the ordination axis median. We considered CWM traits with a Mahalanobis distance greater than the 97.5 \%-quantile of the Chi-square distribution (5.02) to have responded to either lower or higher salinity. Additionally, to test for the effect of different aggregation methods on inferences about CWM trait–salinisation relationships, we repeated the RDA five times, each with family-level aggregated trait values assigned to corresponding taxa from \citet{szocs_effects_2014}.
% Add change in functional diversity

@@

The definitions of grouping features also varied across databases in their differentiation into traits and in their focus, both which can lead to discrepancies in trait definitions (see Table \ref{tab:trait_definitions} and Table \ref{stab:trait_definitions} for a summary of discrepancies in trait definitions). Varying levels of differentiation among the trait databases were present in all investigated grouping features (Table \ref{tab:trait_databases_coding_differentiation}, Table \ref{tab:trait_definitions}, and Table \ref{stab:trait_definitions}).

@@

@@

\section*{Acknowledgements}

@@

% Table 1
\begin{landscape}
  \begin{longtable}{m{2.0cm}|m{4.0cm}|m{4.2cm}|m{4.2cm}|m{1.7cm}|m{2cm}|m{2.2cm}}
      \caption{Excerpt of the comparison of trait definitions between invertebrate trait databases for the traits predator and swimming. The definition is quoted if it enables differences to be identified, otherwise the differences are described. The full version and further information can be found in the supporting information Table \ref{stab:trait_definitions}.}
      \label{tab:trait_definitions}
      \endfirsthead
      \toprule[.1em]
      Trait & \specialcell{Freshwater- \\ ecology.info} & Tachet & CONUS & Vieira & Australia & \specialcell{New \\ Zealand} \\
      \toprule[.1em]
      Feeding predator & 
        "Eating from prey" & 
        \begin{itemize}
            \item Carvers, engulfers \& swallowers
            \item Piercers (plants \& animals) are an additional trait
        \end{itemize} & % Notes: Tachet -> Piercer (plants & animals)
        Engulfers ("ingest prey whole or in parts") \& 
        piercers ("prey tissues and suck fluids") & 
        Predator &
        Piercer \& engulfer &
        Predator
        \\ 
        \midrule
        Locomotion swimming & 
        \begin{itemize}
            \item Passive movement like floating or drifting (trait swimming/scating)
            \item Active movement (trait swimming/diving)
        \end{itemize}. &
        \begin{itemize}
            \item Surface swimmers (over and under the water surface)
            \item Full water swimmers (e.g. Baetidae).
        \end{itemize} & 
        "Adapted for "fishlike" swimming" & 
        Swimmer & 
        Traits swimmer and skater & 
        Swimmers (water column)
        \\
      \bottomrule
  \end{longtable}
\end{landscape}

% Table 2

@@

% Table 5
\begin{landscape}
  \begin{table}[H]
    \centering
    \caption{Percentage of differing cases, minimum, maximum, mean, and standard deviation of absolute differences between trait affinities assigned at family level by experts and aggregated trait affinities from five different aggregation methods.}
    \label{tab:summary_stat_aggr_vs_fam_assigned}
    \begin{tabular}{ll|ccccc}
    \toprule[.1em]
    \specialcell{Data \\ origin} & \specialcell{Comparison to\\ traits at family level} & \specialcell{Differing \\ cases [\%]} & \specialcell{Min. \\ differences} & \specialcell{Max. \\ differences} & \specialcell{Mean abs. \\ differences} & \specialcell{SD abs. \\ differences} \\ 
    \toprule[.1em]
    \multirow{4}{*}{\specialcell{AUS}} & \specialcell{\textit{direct\_agg\textsubscript{median}}} & 16.53 & 0.01 & 1.00 & 0.45 & 0.27 \\ 
    & \specialcell{\textit{direct\_agg\textsubscript{mean}}} & 23.24 & $< 0.01$ & 0.99 & 0.34 & 0.23 \\ 
    & \specialcell{\textit{stepwise\_agg\textsubscript{median}}} & 17.90 & 0.01 & 1.00 & 0.42 & 0.26 \\ 
    & \specialcell{\textit{stepwise\_agg\textsubscript{mean}}} & 23.24 & $< 0.01$ & 0.99 & 0.33 & 0.22 \\ 
    & \specialcell{\textit{weighted\_agg}} & 23.24 & $< 0.01$ & 1.00 & 0.34 & 0.24 \\ 
    \midrule
    \multirow{4}{*}{\specialcell{NA}} & \specialcell{\textit{direct\_agg\textsubscript{median}}} & 15.33 & 0.17 & 1.00 & 0.70 & 0.26 \\ 
    & \specialcell{\textit{direct\_agg\textsubscript{mean}}} & 47.00 & $< 0.01$ & 1.00 & 0.30 & 0.26 \\ 
    & \specialcell{\textit{stepwise\_agg\textsubscript{median}}} & 18.00 & 0.08 & 1.00 & 0.63 & 0.28 \\ 
    & \specialcell{\textit{stepwise\_agg\textsubscript{mean}}} & 47.00 & $< 0.01$ & 1.00 & 0.30 & 0.27 \\ 
    & \specialcell{\textit{weighted\_agg}} & 47.00 & $< 0.01$ & 1.00 & 0.31 & 0.28 \\ 
    \bottomrule
    \end{tabular}
  \end{table}
  \begin{minipage}{\linewidth}{\fontsize{8}{10}\selectfont
    \centering
    Abbreviations: Min., Minimum; Max., Maximum; abs., absolute; SD, Standard deviation; AUS, Australia; NA, North America.
    }
   \end{minipage}
  \end{landscape}

@@

Ranges of aggregated trait affinities for the three examples of phylogenetic structures and simulated levels of trait variability. Shown are the results only for one simulated trait (T1). Similar results were obtained for the other simulated traits (Figure \ref{fig:overview_sim_results_T2_T3}). Boxplots depict results for 100 replicated simulations of each trait aggregation method. The boxplot depicts the median and encompasses the 25\textsuperscript{th} and 75\textsuperscript{th} percentile. Horizontal black lines depict the median. Whiskers extend to the largest and smallest value respectively no further than 1.5 $\times$ the inter-quartile range. Outliers beyond the end of the whiskers are plotted as grey dots. Trait aggregation methods are in order of least to greatest produced ranges to improve visual inspection.
\\

@@

Comparison between the aggregated trait affinities produced by the different trait aggregation methods for every simulated dataset across all 3 simulated traits (5,000 comparisons per scenario). Results are shown for different levels of trait variability which is represented by the standard deviation (SD) of the simulated traits. Dots depict comparisons where absolute differences between aggregated trait affinities were greater than 0.1. Overall, there are 10 possible unique comparisons of which 7 produced absolute differences greater than 0.1. \newline
  Comparisons: \newline
  1) \textit{direct\_agg\textsubscript{median}} - \textit{stepwise\_agg\textsubscript{mean}} \newline
  2) \textit{direct\_agg\textsubscript{median}} - \textit{weighted\_agg} \newline
  3) \textit{stepwise\_agg\textsubscript{mean}} - \textit{stepwise\_agg\textsubscript{median}} \newline
  4) \textit{stepwise\_agg\textsubscript{mean}} - \textit{weighted\_agg} \newline  
  5) \textit{direct\_agg\textsubscript{mean}} - \textit{stepwise\_agg\textsubscript{median}} \newline
  6) \textit{direct\_agg\textsubscript{median}} - \textit{stepwise\_agg\textsubscript{median}} \newline
  7) \textit{stepwise\_agg\textsubscript{median}} - \textit{weighted\_agg}
\\
